Propositions as Sessions, Mechanically

### 2015-04-13 paper/formalisation.tex
Added figures for the sections on GV and the CPS translation.

Following the original work by \citeauthor{Wadler:2014}, the translation from
GV to CP requires defining translations on types and terms of GV, which for
terms is in continuation-passing style.

@@

The translation of from GV types to CP propositons is shown in Figure~\ref{??}
\todo{Coq version or regular version?}. Translation of session types is
noteworthy for taking a GV session type to its notional dual in CP, reflecting
the semantic difference between channels in GV and those in CP.

@@

Due to the locally nameless representation, the translation of GV terms
required type inference of terms to construct corresponding CP terms in order
to specify the propositions. So, the translation takes the environment, in
which the term was typed, as an argument to query the type of free
variables. Unfortunately, this query returns a type of \coqe!option typ!
(\coqe!option! is equivalent to \tm{Maybe} in Haskell) since it may not be
found, but in fact since only well-typed GV terms are translated it is
guaranteed to return something! Therefore, the definition is unnecessarily
cluttered with impossible error cases.

The term translation was initially designed using a monadic approach to
account for the possible error cases in retreiving names from the
environment. The coq-ext-lib
library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
definitions for monad, functor and applicative type classes and associated
instances for basic Coq types. It is similar in style to Haskell's support for
these objects. Unfortunately, this definition did not yield a suitable
function to prove the translation well-typed because the function needs to be
able to operate structurally on not only terms, but also the environment and
the processes being constructed.

An alternative formulation, which avoids the need to handle error cases, is to
first define only the positive cases using an inductive type, the so-called
\textit{graph relation}, and then define a dependent function which only
considers the positive cases within the relation~\cite{McKinna:2009}. However,
several issues were encountered when attempting to specify the
translation. The graph relation requires much of the clutter present in both
the \coqe!cp_rule! and \coqe!wt_tm! definition such as linear enviornments and
cofinite quanitifcation.
\begin{comment}
i'm not sure about that but what is can i say so for example lets see see no i
also have to thing about the general guidelines i can be giving for this
section so we have something to say about guidelines do we yes of course we
have can say that out of this we realised that it is important for a
representation to be good enough to succintly express that which we wish to
say in particular the graph relation the inductive type that we define
propositionally is better one would think than the other way the monadic
possible failure way is not producing a valid thing that we can inductively
induct on so we concern ourself with the proof of translation preserving the
typed term and i think it is important to remember the problem we had with the
inductive hypothesis we had something do with translating the inner term of a
weakening term in a larger context than the one we had available and that was
to do with the destruction of the term inference thing i think no no it was
something else we had the inner term and we had it well typed in an
environment that was smaller sure and then our thing our function was saying
we had some environment that typed both the inner term and the outer weakening
term but we know that is wrong so wrong so here is perhaps where the inductive
type graph relation thing wins out in that we can have a hypothesis for the
inner term which is sufficiently small for the environment and we pattern
match not only on the term but also on the environment yes that is we will do
because it seems to work only when we get to the messy two the messy two
constructor caases will this become a bigg problem inthat we will then have
all this stuff about environment being split and needing to be recombined in
the composite process that is a term not a process but the point still stands
it will need to be about the term and the splitting is really a duplication of
the unfortunately the well typed term relation but otherwise i cannot think
how to reuse or obtain that information for the translation to cp
\end{comment}
Roughly, the theorm aiming to prove is:

@@

\textbf{Modularised development.} Not only is modular development an important
programming practice in general, but in the context of theorem proving it has
the advantage of reducing interactive load time. Executing proof scripts can
be time-consuming especially if those scripts make frequent use of decision
procedures performing proof search, then the execution time is proportional to
the size of the proof context and complexity of the goal. There were cases in
which the system crashed unexpectedly during interactive mode and it is
conceivable that having many large proofs in-memory could be a contributing
factor. For these reasons, the development is partitioned into GV and CP
sections each with sub-divided into definition, infrastructure and typing
modules (CP also has an additional module for reduction lemmata). The CPS
translation is also defined separately. Load time is considerably improved and
the consequences of a crash are mitigated.

\textbf{Representation.} It is important not to go overboard in the
when developing not to use dependent types all the time without working out if
it will be to the advantage of the prover use dependent types and other power
with care for example we could don't want to use mutual inductive definitions
for the types construct since this is overly complicated for what we wish to
achieve and we don't want to index types with kinds since this complicates
further definitions it is about trying to use dependent types responsibly
without using them in an uncritical or thinking manner with great power comes
great responsibility

@@

tactic performance is an issue to be addressed due to the slow proof search a
future work (things like deciding equations on finite sets)

Base rules and rules with binders required permutation assumptions to allow
any ordering of the constituent contexts. a lot of the proofs for well-typed
reductions involved small amounts of permuting environments which were largely
similar across the proofs. while some simple lemmata and tactics where created
to handle the most tedious elements these did not make effective use of
hypotheses from the context. Improved proof search for permutations is
required to simplify such developments.

>>>>

Following the original work by \citeauthor{Wadler:2014}, the translation from
GV to CP requires defining translations on types and terms of GV.

@@

\input{gv-trans-types}

The translation of from GV types to CP propositons is shown in
Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
taking a GV session type to its notional dual in CP, reflecting the semantic
difference between channels in GV and those in CP. For GV, the type refers to
the action on the channel endpoint. For CP, the proposition refers to the
\textit{interface} exposed to a user of the channel. For example, output in GV
corresponds to input in CP; the action on the channel endpoint is to output a
value, and to do this the channel must input the value.

@@

\input{gv-cps-cp}

The translation of GV terms to CP processes is given, following
\citeauthor{Wadler:2014}, using continuation-passing style. In this style a
well-typed term in GV is parameterised by a name representing the channel
which communicates the result of the translation (a CP process). Due to the
locally nameless representation, the translation of GV terms required a
procedure to infer the type of terms in order to specify the binding type for
the corresponding CP process. An example, translating choice, is shown in
Figure~\ref{fig:trans}, note that I am using \citeauthor{Wadler:2014}'s style
but using an LNCQ notation for clarity. In order to construct the cut and
output processes, the type of \tm{M} and \tm{N} must be computed. So, the
translation takes the environment, in which the term is typed, as an argument
to query the type of free variables. Unfortunately, this query returns a type
of \coqe!option typ!  (\coqe!option! is equivalent to \tm{Maybe} in Haskell)
since it may not be found, but in fact since only well-typed GV terms are
translated it is guaranteed to return something!  Therefore, the definition is
unnecessarily cluttered with impossible error cases.

The term translation was initially designed using a monadic approach to
account for the possible error cases in retreiving names from the
environment. The coq-ext-lib
library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
definitions for monad, functor and applicative type classes and associated
instances for basic Coq types. It is similar in style to Haskell's support for
these objects. Unfortunately, this definition did not yield a suitable
function to prove the translation well-typed because the function needs to be
able to operate structurally on not only terms, but also the environment and
the processes being constructed. Consider again the translation of case. here
we require to consider nl and nr in a separate environment than m and type the
resulting case term in the combined environment we also need to open these
terms using cofinite quantification to yield a suitable induction
principle. all told the function is not readily definable. further problems do
arise as we shall see.

An alternative formulation, which avoids the need to handle error cases, is to
first define only the positive cases using an inductive type, the so-called
\textit{graph relation}, and then define a dependent function that only
considers the positive cases for which the relation is
defined~\cite{McKinna:2009}. However, several issues were encountered when
attempting to specify the translation. The graph relation requires much of the
clutter present in both the \coqe!cp_rule! and \coqe!wt_tm! definition such as
linear enviornments and cofinite quantification. the proof of cps translation
using the graph relation is a work in progress. the main difficulty is that
the representation requires opening both GV terms and CP processes in
hypotheses which would appear to duplicate a lot of the overhead of their
respective rule definitions.  again, representation is important even if the
proof has still eluded us.  while it is remains to be seen whether or not the
graph relation assists us in these matters it is nonetheless worth trying to
express a property in multiple ways

Roughly, the theorm aiming to prove is:
\begin{coq}
Theorem cps_trans_wt:
  forall $\Phi$ M T z P $\Gamma$ $\GD$
         (WT: $\Phi$ $\tpgv$ M$\of{T}$)
         (NIN: z $\notin$ GVFV M $\cup$ dom $\Phi$)
         (ENV: trans_env $\Phi$ = $\GD$)
         (PER: Permutation $\Gamma$ ($\GD$ ++ z $\sim$ $\sem{T}$t))
         (CP: transTm M $\Phi$ z P),
    P $\tpcp$ $\Gamma$.
\end{coq}

@@

\textbf{Modularised development.} Not only is modular development an important
programming practice in general, but in the context of theorem proving it has
the advantage of reducing interactive load time. Executing proof scripts can
be time-consuming especially if those scripts make frequent use of decision
procedures performing proof search, then the execution time is proportional to
the size of the proof context and complexity of the goal. There were cases in
which the system crashed unexpectedly during interactive mode and it is
conceivable that having many large proofs in-memory could be a contributing
factor. For these reasons, the development is partitioned into GV and CP
sections each further sub-divided into definition, infrastructure and typing
modules (CP also has an additional module for reduction lemmata). The CPS
translation is also defined separately. Load time is considerably improved and
the consequences of a crash are mitigated.

\textbf{Representation.} As discussed in the previous sections representation
choice is a central issue in formalising programming language
metatheory. Decisions on how to define, for instance, types, calculi and
translations impact on future definitions and properties dependent on
them. Considerations range from exploiting the full expressivity of the
theorem prover by using dependent types in the case of Coq. to how to encode
the properties of interest and being able to assess accurately the most
fruitful encoding upfront. avoiding features that would provide a more faitful
relationship between formal and informal systems is desirable if those
features would unnecessarily complicated the reasoning. it is a tradeoff.

@@

\textbf{Effective Automation.} It is worth considering situations where
automation would simplify proofs. Automation is already an important
consideration in the interactive theorem proving community. Base rules and
rules with binders required permutation assumptions to allow any ordering of
the constituent contexts. a lot of the proofs for well-typed reductions
involved small amounts of permuting environments which were largely similar
across the proofs. while some simple lemmata and tactics where created to
handle the most tedious elements these did not make effective use of
hypotheses from the context. Improved proof search for permutations is
required to simplify such developments.
